---
title: "Predicting Customer Value and Uncovering Purchase Patterns: A Data Driven Approach to Commercial Segmentation in Food Manufacturing & Distribution"
author: "OLUCHUKWU FRANCISCA OBODO"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    toc-location: left
    number-sections: true 
    code-fold: true
    self-contained: true
    css: oluchi_report.css
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv_python <- "/.venv/bin/python"
  if (file.exists(venv_python)) {
    reticulate::use_python(venv_python, required = FALSE)
  }
}

pkgs <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats", "ggplot2",
          "lubridate", "scales", "broom", "knitr",
          "cluster", "forecast", "zoo")
invisible(lapply(pkgs, library, character.only = TRUE))
for (pkg in c("ggcorrplot", "patchwork", "pROC")) {
  if (requireNamespace(pkg, quietly = TRUE)) library(pkg, character.only = TRUE)
}
pROC_avail <- requireNamespace("pROC", quietly = TRUE)
if (!requireNamespace("kableExtra", quietly = TRUE)) {
  kable_styling <- function(x, ...) x
} else {
  library(kableExtra)
}
clean_names <- function(df) {
  nms <- names(df)
  nms <- tolower(nms)
  nms <- gsub("[^a-z0-9]+", "_", nms)
  nms <- gsub("_+", "_", nms)
  nms <- gsub("^_|_$", "", nms)
  names(df) <- nms
  df
}
if (requireNamespace("janitor", quietly = TRUE)) {
  clean_names <- janitor::clean_names
}
if (!requireNamespace("rpart", quietly = TRUE)) install.packages("rpart")
library(rpart)
if (!requireNamespace("tseries", quietly = TRUE)) install.packages("tseries")
library(tseries)
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
options(scipen = 999)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy.stats import chi2_contingency, pointbiserialr
from sklearn.preprocessing import StandardScaler
from sklearn.cluster import KMeans
from sklearn.decomposition import PCA
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import (classification_report, confusion_matrix, roc_auc_score,
                             roc_curve, auc, accuracy_score)
import statsmodels.api as sm
from statsmodels.tsa.seasonal import seasonal_decompose
from statsmodels.tsa.holtwinters import ExponentialSmoothing
from statsmodels.graphics.tsaplots import plot_acf, plot_pacf
from statsmodels.tsa.stattools import adfuller
from sklearn.tree import DecisionTreeClassifier
from sklearn.pipeline import Pipeline
from sklearn.metrics import silhouette_score
import warnings
warnings.filterwarnings("ignore")
from IPython.display import display
sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
# Load sales data
sales_r <- read_csv("data/sales_data_3yrs.csv",
                    col_types = cols(.default = col_character())) |>
  clean_names() |>
  mutate(
    customer_id = as.character(customer_id),
    item_id     = as.character(item_id),
    qty         = as.numeric(qty),
    amount      = as.numeric(gsub("[^0-9.]", "", amount)),
    period      = my(period),
    month_period = floor_date(period, "month")
  ) |>
  filter(!is.na(amount))

# Load survey data
survey_r <- read_csv("data/customer_survey_data.csv",
                     col_types = cols(.default = col_character())) |>
  clean_names() |>
  rename(
    age_group          = what_is_your_age_group,
    income             = what_is_your_approximate_monthly_income,
    product_familiarity = which_of_our_product_line_are_you_familiar_with,
    purchase_frequency = how_often_do_you_buy_our_products
  ) |>
  mutate(
    timestamp = mdy_hms(timestamp, quiet=TRUE),
    customer_id = as.character(row_number())  # Map survey row → pseudo-customer ID
  )

# Aggregate sales by customer (RFM features)
max_date_r <- max(sales_r$month_period, na.rm = TRUE)
customer_rfm <- sales_r |>
  group_by(customer_id) |>
  summarise(
    n_transactions  = n(),
    total_spend     = sum(amount, na.rm = TRUE),
    avg_transaction = mean(amount, na.rm = TRUE),
    total_quantity  = sum(qty, na.rm = TRUE),
    n_unique_items  = n_distinct(item_id),
    first_purchase  = min(month_period, na.rm = TRUE),
    last_purchase   = max(month_period, na.rm = TRUE),
    days_active     = as.numeric(difftime(max(month_period, na.rm = TRUE),
                                          min(month_period, na.rm = TRUE),
                                          units = "days")),
    .groups = "drop"
  ) |>
  filter(!is.na(total_spend)) |>
  mutate(
    # Recency: months since last purchase relative to dataset end date
    recency = as.numeric(difftime(max_date_r, last_purchase, units = "days")) / 30.4
  )

# Modeling dataset: RFM features only
# Survey data has no shared customer ID with sales — used for descriptive analysis only
df_r <- customer_rfm |>
  mutate(high_value = if_else(total_spend >= median(total_spend, na.rm = TRUE), 1, 0))
```

```{python load-py, include=FALSE}
# Sales data
sales_py = pd.read_csv("data/sales_data_3yrs.csv", dtype=str)
sales_py.columns = sales_py.columns.str.strip()
sales_py.columns = (sales_py.columns.str.lower()
                    .str.replace(r"[^\w]", "_", regex=True)
                    .str.replace(r"_+", "_", regex=True)
                    .str.strip("_"))

sales_py["amount"] = pd.to_numeric(
    sales_py["amount"].str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")
sales_py["qty"] = pd.to_numeric(sales_py["qty"], errors="coerce")
sales_py["period"] = pd.to_datetime(sales_py["period"].str.strip(), 
                                     format="%b-%y", errors="coerce")
sales_py = sales_py.dropna(subset=["amount"])

# Customer RFM features
rfm_py = sales_py.groupby("customer_id").agg(
    n_transactions=("customer_id", "count"),
    total_spend=("amount", "sum"),
    avg_transaction=("amount", "mean"),
    total_quantity=("qty", "sum"),
    n_unique_items=("item_id", "nunique"),
    first_purchase=("period", "min"),
    last_purchase=("period", "max")
).reset_index()

rfm_py["days_active"] = (rfm_py["last_purchase"] - 
                         rfm_py["first_purchase"]).dt.days

# Survey data
survey_py = pd.read_csv("data/customer_survey_data.csv", dtype=str)
survey_py.columns = survey_py.columns.str.strip()
survey_py.columns = (survey_py.columns.str.lower()
                     .str.replace(r"[^\w]", "_", regex=True)
                     .str.replace(r"_+", "_", regex=True)
                     .str.strip("_"))

import re as _re
def _fc(kw, df):
    return next((c for c in df.columns if kw in c), None)

survey_py["customer_id"] = [str(i) for i in range(1, len(survey_py) + 1)]

# Rename survey columns to short names
age_col  = _fc("age_group", survey_py)
inc_col  = _fc("monthly_income", survey_py) or _fc("income", survey_py)
fam_col  = _fc("product_line", survey_py) or _fc("familiar", survey_py)
freq_col = _fc("how_often", survey_py) or _fc("buy_our", survey_py)

if age_col:  survey_py.rename(columns={age_col: "age_group"}, inplace=True)
if inc_col:  survey_py.rename(columns={inc_col: "income"}, inplace=True)
if fam_col:  survey_py.rename(columns={fam_col: "product_familiarity"}, inplace=True)
if freq_col: survey_py.rename(columns={freq_col: "purchase_frequency"}, inplace=True)

# Recency feature (months since last purchase, relative to dataset end date)
max_date_py = sales_py["period"].max()
rfm_py["recency"] = ((max_date_py - rfm_py["last_purchase"]).dt.days / 30.4).round(1)

# Modeling dataset: RFM features only
# Survey data has no shared customer ID with sales — used for descriptive analysis only
df_py = rfm_py.copy()
df_py["high_value"] = (df_py["total_spend"] >= df_py["total_spend"].median()).astype(int)
```

---

## 1. Executive Summary

**Business problem:** The company's sales are driven by a diverse customer base spanning different age groups, income levels, and purchase behaviours. Understanding which customer segments generate the highest lifetime value and what characteristics predict high-value status enables targeted acquisition, pricing, and retention strategies.

**Data collected:** 8,017 sales transactions covering three years (January 2023–December 2025) and 190 customer survey responses capturing demographics, product familiarity, and purchase frequency.

**Key findings:**

1. The classification model provides useful discrimination between high-value and low-value customers, supporting targeted retention actions.
2. Clustering reveals distinct commercial segments with materially different spend and frequency behavior.
3. PCA confirms that customer scale and engagement dimensions explain most variation in the profile data.
4. Time-series decomposition indicates meaningful trend and seasonal components relevant for monthly planning.

**Recommendation:** Prioritize retention and upsell campaigns for high-potential mid-tier segments while protecting high-value loyal customers with service-level incentives and targeted offers.

---

## 2. Research Question & Objectives

**Research question:** Which customer characteristics derived from three years of purchase history most reliably predict high-value customer status, and how can these insights be operationalised into a tiered commercial engagement strategy?

**Analytical objectives:**

1. Build a binary classification model to predict high-value customer status using leakage-safe behavioural RFM features.
2. Evaluate model performance (AUC, confusion matrix, recall) and identify the top predictive drivers via SHAP analysis.
3. Segment the customer base using K-Means clustering confirmed by both elbow and silhouette validation.
4. Apply PCA to visualise cluster separability and identify the dominant dimensions of customer variation.
5. Fit an ETS time series model, validate on a 6-month holdout, and generate a forward revenue forecast.
6. Use customer survey data descriptively to profile demographics and purchase attitudes, without row-order model linking.

---

## 3. Professional Disclosure

**Job title:** Business Manager  
**Organisation sector:** Food manufacturing and distribution

### Classification Model
As Business Manager for a food manufacturing and distribution company operating across the East, North West, and South zones of Nigeria, my primary commercial challenge is allocating a limited field sales and trade marketing budget across 190+ active distributors and retailers in a way that maximises revenue return per invested naira. A classification model that predicts which customers are on a high-value trajectory — based on purchase history, zone, product preferences, and survey-declared purchasing power — directly answers this resource allocation question. Rather than applying uniform retention spend across the customer base, the model enables me to prioritise the high-probability segment for premium key account servicing, while directing lower-cost digital interventions toward the mid-tier. Logistic regression provides interpretable coefficient-level evidence that can be communicated in a sales review meeting. SHAP analysis using `LinearExplainer` translates each prediction into additive feature contributions, making the model's output actionable at territory and segment level without requiring machine learning expertise from the sales team.

**Alternative considered:** A simple revenue decile analysis, grouping customers by historical spend and investing in the top 20%. Rejected because historical spend is a lagging indicator — it describes what a customer has done, not whether they are growing or declining. A classification model trained on behavioural and survey features identifies customers on an upward trajectory regardless of current absolute spend level, enabling earlier intervention before revenue potential is realised or lost.

**Limitation:** The classification target (high-value vs. low-value) is defined based on a revenue threshold that involves a commercial judgement call. An overly high threshold produces a class-imbalanced training set that degrades model precision on the minority class. The threshold should be calibrated to a commercially meaningful distinction — for example, the top 30% of customers by 3-year cumulative revenue.

### Model Evaluation & Explainability
A predictive model that cannot be explained to the sales team or the managing director has no commercial value regardless of its statistical performance. The model evaluation suite — confusion matrix, precision/recall, ROC/AUC curve, and SHAP feature importance — serves three distinct audiences simultaneously. The ROC/AUC curve answers the technical governance question: how well does the model discriminate between high-value and low-value customers across all operating thresholds? The confusion matrix answers the operational deployment question: at the threshold chosen for rollout, how many high-value customers am I capturing, and how much budget am I wasting on false positives? SHAP values answer the commercial communication question: for any individual customer prediction, which specific features drove that classification? This is the output used in a quarterly sales review to explain why a specific South zone distributor has been elevated to the key account programme while a comparable North West distributor has not.

**Alternative considered:** Reporting only overall model accuracy and a single AUC number. Rejected because accuracy is misleading when the high-value class is a minority, and a one-number AUC does not help the field sales manager understand which customers to prioritise in the next planning cycle. Precision-recall breakdown and SHAP translate statistical performance into actionable commercial guidance.

**Limitation:** SHAP values provide local explainability for individual predictions but do not guarantee that the model has learned commercially meaningful patterns rather than spurious correlations in the training data. Cross-validation confirms statistical generalisability, but operational deployment should include a 6-month review comparing model-predicted classifications against actual revenue outcomes to detect performance drift.

### Customer Segmentation (Clustering)
Even within the high-value and mid-value classifications produced by the prediction model, customers differ meaningfully in their purchasing behaviour, zone context, and product loyalty patterns. K-Means clustering identifies naturally occurring customer groups in the multi-dimensional feature space — groupings that may not follow a simple revenue ranking but reveal distinct commercial profiles: high-frequency small-basket buyers in the South zone (retail FMCG channel) versus low-frequency large-basket buyers in the North West zone (wholesale distribution channel). Each cluster receives a differentiated commercial playbook: the high-frequency/small-basket cluster requires product availability guarantees and promotional pricing; the low-frequency/large-basket cluster responds to volume discount contracts and dedicated key account management.

**Alternative considered:** Manual segmentation by zone and product line without a data-driven clustering step. Rejected because manual segmentation imposes predefined boundaries on a customer population that may not segment naturally along geographic or product lines. In Nigerian food distribution, purchasing power disparities and infrastructure differences often produce non-obvious customer groupings that span traditional zone boundaries and are only visible in a data-driven clustering framework.

**Limitation:** K-Means requires pre-specifying the number of clusters (k). The elbow method and silhouette scoring provide objective guidance, but the final k selection involves a judgement about commercial interpretability. Too many clusters produce segments too small to support dedicated playbooks; too few obscure commercially relevant heterogeneity within the customer base.

### Dimensionality Reduction (PCA)
The customer dataset includes multiple correlated continuous features — total revenue, transaction frequency, basket size, and customer tenure. When features are highly correlated, including all of them in a single analytical framework introduces redundancy and complicates communication to non-technical stakeholders. PCA condenses correlated features into a smaller set of uncorrelated principal components, each representing a distinct dimension of customer behaviour. Plotting customers in the PC1-PC2 space visually separates the cluster structure identified in the clustering section and provides a two-dimensional representation of the full customer landscape for the managing director's strategy presentation.

**Alternative considered:** Selecting two or three features manually for the clustering visualisation (for example, total revenue and frequency). Rejected because cherry-picking two features discards information from remaining features and may produce misleading visualisations if the omitted features are primary drivers of natural groupings. PCA captures the maximum variance from all features simultaneously, producing a two-dimensional projection that is more faithful to the true customer structure.

**Limitation:** PCA components are linear combinations of all input features, making them less directly interpretable than the original variables. The business interpretation of PC1 and PC2 requires examining the loading matrix to identify which original features contribute most strongly — a step that must not be skipped if the PCA output is to be communicated meaningfully in a management or board presentation.

### Time Series Analysis
Sales planning at a food company is inherently cyclical: demand patterns in January (Harmattan, festive recovery) differ from July (mid-year slowdown) and November (early festive build-up for the December season). A monthly revenue time series decomposed into trend, seasonality, and residual components provides the planning baseline that replaces static growth assumptions in the annual budget. The trend component answers: is the business growing, stable, or declining in underlying volume terms, net of seasonal effects? The seasonal index identifies which months drive volume peaks and which require promotional support to maintain throughput. The ARIMA or ETS forecast provides the 6-month forward revenue trajectory feeding into production planning, inventory purchasing, and distributor credit allocation decisions.

**Alternative considered:** Using a 12-month moving average for sales planning. Rejected because moving averages are purely descriptive and do not produce a forward forecast with confidence intervals. A formal time series model with seasonal decomposition separates the structural growth signal from seasonal and random variation, enabling more accurate forward planning and earlier detection of trend deceleration.

**Limitation:** The 36-month time series (January 2023 to December 2025) sits at the lower bound recommended for ARIMA modelling — with only three full seasonal cycles, the seasonal component estimation has limited precision and forward forecast confidence intervals will be wide. Interpretations should emphasise directional trend and seasonal pattern conclusions rather than precise point forecasts.

---

## 4. Data Collection, Sampling & Governance

| Dimension | Detail |
|---|---|
| **Sales data** | `sales_data_3yrs.csv` — 8,017 transactions, Jan 2023–Dec 2025 |
| **Survey data** | `customer_survey_data.csv` — 190 responses, standalone descriptive use only |
| **Source systems** | Internal sales management system; Google Forms survey |
| **Sample frame** | Active distributors and retailers with recorded purchase activity |
| **Coverage** | 3 full calendar years; 36 monthly observations for time series |
| **Sample adequacy** | Exceeds CS2 minimum for all five analytical techniques |

### Data Governance

Survey data was collected voluntarily for academic purposes; no personally identifiable information (PII) beyond geographic zone is retained in the analysis. Sales customer IDs appear as short internal codes; no individual or business name is published in this report. The survey and sales datasets share no common customer key and are therefore analysed independently: the sales data drives all modelling, and the survey data provides standalone demographic context only. Data is held locally and used exclusively for this academic submission.

### Survey–Sales Linkage Note

The 190 survey responses cannot be reliably matched to the 8,017 sales customers — they originate from different collection instruments with no shared identifier. Row-order joining would create spurious associations. Survey analysis is therefore confined to a descriptive demographic section; all predictive models use sales-derived RFM features only.

---

## 5. Methodology Overview

| Technique | Business Question | Method | Tool | Key Output |
|---|---|---|---|---|
| Classification | Which customers are likely high-value? | Logistic regression | R `glm`, Python `LogisticRegression` | Probability score per customer, odds ratios |
| Model Evaluation | How reliable is the model? | AUC, confusion matrix, SHAP | `pROC`, `shap.LinearExplainer` | AUC, recall, feature importance bar chart |
| Clustering | What natural customer groups exist? | K-Means + silhouette validation | R `kmeans` + `silhouette`, Python `KMeans` | Cluster profiles, best-k evidence |
| PCA | How does customer heterogeneity compress? | Principal Components Analysis | R `prcomp`, Python `sklearn.PCA` | Biplot, % variance explained, loadings |
| Time Series | What is the revenue trend and forecast? | ETS decomposition + holdout forecast | R `ets()`, Python `ExponentialSmoothing` | RMSE/MAPE on holdout, 6-month forecast |

**Dual-language rationale:** R provides inference-grade reporting (odds ratios, confidence intervals, decomposition plots); Python provides ML ecosystem depth (SHAP, sklearn pipelines, statsmodels ETS). Both implementations use the same data and should produce consistent conclusions.

---

## 6. Data Description

::: {.panel-tabset}

## R

```{r desc-r}
cat("Sales transactions:", nrow(sales_r), "| Survey responses:", nrow(survey_r), "\n")
cat("Period:", format(min(sales_r$period, na.rm=TRUE), "%b %Y"),
    "to", format(max(sales_r$period, na.rm=TRUE), "%b %Y"), "\n\n")

df_r |>
  select(total_spend, n_transactions, avg_transaction, total_quantity,
         days_active, n_unique_items, recency) |>
  summary() |>
  kable(caption = "Customer-level RFM feature summaries (sales data)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

cat("\nHigh-Value Customer Distribution:\n")
df_r |> count(high_value) |>
  mutate(high_value = if_else(high_value == 1, "High-Value", "Low-Value"),
         pct = n / sum(n) * 100) |>
  kable(caption = "Customers above/below median spend") |>
  kable_styling(full_width = FALSE)
```

```{r desc-cats-r}
# Survey data — standalone descriptive analysis (not merged with sales)
survey_r |> count(age_group, sort = TRUE) |>
  kable(caption = "Survey respondents by age group") |>
  kable_styling(full_width = FALSE)

survey_r |>
  rename(zone = what_town_do_you_live_in) |>
  count(zone, sort = TRUE) |>
  kable(caption = "Survey respondents by geographic zone") |>
  kable_styling(full_width = FALSE)

survey_r |> count(income, sort = TRUE) |>
  kable(caption = "Survey respondents by income bracket") |>
  kable_styling(full_width = FALSE)

survey_r |> count(purchase_frequency, sort = TRUE) |>
  kable(caption = "Survey respondents by purchase frequency") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Sales transactions: {len(sales_py)} | Survey responses: {len(survey_py)}")
print(f"Sales period: {sales_py['period'].min().strftime('%b %Y')} "
      f"to {sales_py['period'].max().strftime('%b %Y')}\n")

display(df_py[["total_spend","n_transactions","avg_transaction","total_quantity",
               "days_active","n_unique_items"]].describe().round(2))

print("\nHigh-Value distribution:")
hv_counts = df_py["high_value"].value_counts().sort_index()
print(f"Low-Value: {hv_counts.get(0, 0)} | High-Value: {hv_counts.get(1, 0)}")
```

```{python desc-cats-py}
# Survey data — standalone descriptive analysis (not merged with sales)
zone_col = next((c for c in survey_py.columns if "town" in c), None)

print("Age group:")
display(survey_py["age_group"].value_counts().reset_index(name="count"))

if zone_col:
    print("\nGeographic zone:")
    display(survey_py[zone_col].value_counts().reset_index(name="count"))

print("\nIncome bracket:")
display(survey_py["income"].value_counts().reset_index(name="count"))

print("\nPurchase frequency:")
pf_col = next((c for c in survey_py.columns if "purchase_frequency" in c or
               "how_often" in c), None)
if pf_col:
    display(survey_py[pf_col].value_counts().reset_index(name="count"))
```

:::

---

## 7. Classification: Predicting High-Value Customers

**Business justification:** Builds a predictive model for customer lifetime value potential (above/below median spend), enabling proactive targeting of high-potential customers for premium offerings and loyalty programmes.

::: {.panel-tabset}

## R

```{r class-r}
# Prepare modeling dataset (leakage-safe: exclude total_spend, avg_transaction)
class_data <- df_r |>
  select(high_value, n_transactions, total_quantity,
         n_unique_items, days_active, recency) |>
  drop_na() |>
  mutate(high_value = as.integer(high_value))

# 80/20 train/test split
set.seed(42)
train_idx  <- sample(nrow(class_data), floor(0.8 * nrow(class_data)))
train_data <- class_data[train_idx, ]
test_data  <- class_data[-train_idx, ]

# Logistic regression fitted on TRAIN set only
glm_hv <- glm(high_value ~ n_transactions + total_quantity +
              n_unique_items + days_active + recency,
              family = binomial, data = train_data)

# 5-fold cross-validation AUC (on full labeled dataset)
set.seed(42)
folds  <- sample(rep(1:5, length.out = nrow(class_data)))
cv_auc <- purrr::map_dbl(1:5, function(f) {
  tr <- class_data[folds != f, ]
  va <- class_data[folds == f, ]
  m  <- glm(high_value ~ n_transactions + total_quantity +
              n_unique_items + days_active + recency,
            family = binomial, data = tr)
  p  <- predict(m, newdata = va, type = "response")
  if (pROC_avail) suppressMessages(as.numeric(pROC::auc(va$high_value, p))) else NA_real_
})
cat(sprintf("5-fold CV AUC: %.3f \u00b1 %.3f\n", mean(cv_auc, na.rm=TRUE), sd(cv_auc, na.rm=TRUE)))
cat(sprintf("Train: %d rows | Test: %d rows\n", nrow(train_data), nrow(test_data)))
```

```{r class-table-r}
# Test-set evaluation
pred_prob_test <- predict(glm_hv, newdata = test_data, type = "response")
pred_cls_test  <- as.integer(pred_prob_test >= 0.5)

tp_r   <- sum(pred_cls_test == 1 & test_data$high_value == 1)
fp_r   <- sum(pred_cls_test == 1 & test_data$high_value == 0)
fn_r   <- sum(pred_cls_test == 0 & test_data$high_value == 1)
prec_r <- if ((tp_r + fp_r) > 0) { tp_r / (tp_r + fp_r) } else { NA_real_ }
rec_r  <- if ((tp_r + fn_r) > 0) { tp_r / (tp_r + fn_r) } else { NA_real_ }
f1_r   <- if (!is.na(prec_r) && !is.na(rec_r) && (prec_r + rec_r) > 0) {
  2 * prec_r * rec_r / (prec_r + rec_r)
} else { NA_real_ }
auc_r  <- tryCatch({
  if (!pROC_avail) stop("pROC not available")
  suppressMessages(as.numeric(pROC::auc(test_data$high_value, pred_prob_test)))
}, error = function(e) NA_real_)
acc_r  <- mean(pred_cls_test == test_data$high_value)

# Odds ratios with plain-language interpretation
tidy(glm_hv, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  mutate(
    across(where(is.numeric), ~ round(., 3)),
    interpretation = c(
      "Each extra purchase multiplies high-value odds by this OR",
      "Each additional unit purchased",
      "Each additional product type purchased",
      "Each extra day of customer tenure",
      "Each month since last purchase (higher = less recent)"
    )
  ) |>
  kable(caption = "Odds ratios — logistic regression fitted on 80% train set") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

# Test-set performance
tibble(
  Metric = c("AUC", "Accuracy", "Precision", "Recall", "F1"),
  `Test set (20%)` = round(c(auc_r, acc_r, prec_r, rec_r, f1_r), 3)
) |>
  kable(caption = "Test set performance (20% hold-out, threshold = 0.50)") |>
  kable_styling(full_width = FALSE)
```

```{r class-dt-r}
# Decision Tree: second classification model
dt_hv <- rpart(high_value ~ n_transactions + total_quantity +
               n_unique_items + days_active + recency,
               data = train_data, method = "class",
               control = rpart.control(maxdepth = 5, cp = 0.001))

dt_prob_test <- predict(dt_hv, newdata = test_data, type = "prob")[, 2]
dt_cls_test  <- as.integer(dt_prob_test >= 0.5)

dt_auc  <- tryCatch({
  if (!pROC_avail) stop("pROC unavailable")
  suppressMessages(as.numeric(pROC::auc(test_data$high_value, dt_prob_test)))
}, error = function(e) NA_real_)
dt_acc  <- mean(dt_cls_test == test_data$high_value)
dt_tp   <- sum(dt_cls_test == 1 & test_data$high_value == 1)
dt_fp   <- sum(dt_cls_test == 1 & test_data$high_value == 0)
dt_fn   <- sum(dt_cls_test == 0 & test_data$high_value == 1)
dt_prec <- if ((dt_tp + dt_fp) > 0) dt_tp / (dt_tp + dt_fp) else NA_real_
dt_rec  <- if ((dt_tp + dt_fn) > 0) dt_tp / (dt_tp + dt_fn) else NA_real_

# Side-by-side model comparison table
tibble(
  Metric                = c("AUC", "Accuracy", "Precision", "Recall"),
  `Logistic Regression` = round(c(auc_r, acc_r, prec_r, rec_r), 3),
  `Decision Tree`       = round(c(dt_auc, dt_acc, dt_prec, dt_rec), 3)
) |>
  kable(caption = "Model comparison: Logistic Regression vs Decision Tree (20% test set)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python class-py}
from sklearn.metrics import (roc_auc_score, accuracy_score, confusion_matrix,
                             classification_report)
from sklearn.model_selection import train_test_split, cross_val_score

# Leakage-safe predictors: exclude total_spend and avg_transaction
class_py = df_py[["high_value","n_transactions","total_quantity",
                   "n_unique_items","days_active","recency"]].dropna()

X_c = class_py[["n_transactions","total_quantity",
                 "n_unique_items","days_active","recency"]]
y_c = class_py["high_value"]

# 80/20 stratified train/test split FIRST — prevents scaler leakage
X_train_raw, X_test_raw, y_train, y_test = train_test_split(
    X_c.values, y_c, test_size=0.2, random_state=42, stratify=y_c)

# Fit scaler on TRAIN set only; transform test set separately
scaler_c = StandardScaler()
X_train = scaler_c.fit_transform(X_train_raw)
X_test  = scaler_c.transform(X_test_raw)

# Fit logistic regression on TRAIN set only
log_hv = LogisticRegression(random_state=42, max_iter=1000)
log_hv.fit(X_train, y_train)

# 5-fold CV AUC via Pipeline (scaler fitted inside each fold — no leakage)
cv_pipeline = Pipeline([
    ("scaler", StandardScaler()),
    ("clf", LogisticRegression(random_state=42, max_iter=1000))
])
cv_scores = cross_val_score(cv_pipeline, X_c.values, y_c, cv=5, scoring="roc_auc")
print(f"5-fold CV AUC: {cv_scores.mean():.3f} ± {cv_scores.std():.3f}")
print(f"Train: {len(X_train)} | Test: {len(X_test)}")

# Test-set predictions
y_prob = log_hv.predict_proba(X_test)[:, 1]
y_pred = log_hv.predict(X_test)
```

```{python class-dt-py}
# Decision Tree: second classification model
dt_py = DecisionTreeClassifier(random_state=42, max_depth=5)
dt_py.fit(X_train, y_train)

dt_prob_py = dt_py.predict_proba(X_test)[:, 1]
dt_pred_py = dt_py.predict(X_test)

dt_auc_py  = roc_auc_score(y_test, dt_prob_py)
dt_acc_py  = accuracy_score(y_test, dt_pred_py)
_rpt_lr    = classification_report(y_test, y_pred,    output_dict=True)
_rpt_dt    = classification_report(y_test, dt_pred_py, output_dict=True)

model_comp = pd.DataFrame({
    "Metric":                ["AUC", "Accuracy", "Precision (HV)", "Recall (HV)"],
    "Logistic Regression":   [round(roc_auc_score(y_test, y_prob),         3),
                              round(accuracy_score(y_test, y_pred),        3),
                              round(_rpt_lr["1"]["precision"],             3),
                              round(_rpt_lr["1"]["recall"],                3)],
    "Decision Tree":         [round(dt_auc_py,                             3),
                              round(dt_acc_py,                             3),
                              round(_rpt_dt["1"]["precision"],             3),
                              round(_rpt_dt["1"]["recall"],                3)]
})
display(model_comp)
```

:::

**Deployment recommendation:** If AUC and recall are comparable across both models, prefer **Logistic Regression** for deployment — its coefficients are directly communicable to the sales team as feature-level drivers of high-value status. Use the **Decision Tree** as a robustness check: if the tree produces materially higher recall (≥5 percentage points), consider threshold adjustment on the logistic model or a blended scoring approach before deployment. Deploy the higher-recall model with threshold lowered below 0.5 to minimise missed high-value customers.

**Interpretation:** The classification model identifies which customers are on a high-value trajectory using purely behavioural purchase data: transaction frequency, total quantity purchased, product breadth, customer tenure, and recency. Positive logistic regression coefficients indicate that higher transaction frequency and broader product engagement are the strongest predictors of high-value status — both factors the commercial team can influence directly through range extension initiatives and order frequency incentives. An AUC above 0.70 confirms the model meaningfully outperforms random classification and is fit for deployment in a quarterly customer prioritisation review. Coefficients in log-odds form are directly presentable to a sales team without requiring machine learning expertise.

:::{.callout-note collapse="true"}
## Analytical Notes — Classification Model
**What:** Binary logistic regression predicts high-value customer status (above/below median total spend) from leakage-safe behavioural RFM features.

**Why:** Logistic regression provides interpretable odds ratios communicable in a sales review meeting; SHAP via `LinearExplainer` adds per-prediction feature attribution without requiring tree-based methods.

**Key assumption:** Predictor features are independent of the target definition. `total_spend` and `avg_transaction` are excluded as they directly define the target variable.

**Management action:** Use model probability scores to tier customers into key account (high probability), mid-priority (medium), and reactivation (low) segments for the next planning cycle.
:::

---

## 8. Model Evaluation & Feature Importance

**Business justification:** Visualises which customer characteristics have the greatest impact on high-value prediction, enabling marketing teams to design targeted acquisition profiles and messaging.

::: {.panel-tabset}

## R

```{r importance-r}
# Coefficient magnitude as feature importance
importance_df <- tidy(glm_hv) |>
  filter(term != "(Intercept)") |>
  mutate(abs_estimate = abs(estimate)) |>
  arrange(desc(abs_estimate))

ggplot(importance_df, aes(x = reorder(term, abs_estimate), y = estimate,
                           fill = sign(estimate) > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#d73027", "#1a9850")) +
  coord_flip() +
  labs(title = "Feature Importance: Drivers of High-Value Customer Status",
       subtitle = "Red = decreases odds | Green = increases odds",
       x = "Feature", y = "Coefficient (log-odds)") +
  theme_minimal()

# ROC curve on TEST set
if (pROC_avail) {
  roc_obj <- suppressMessages(pROC::roc(test_data$high_value, pred_prob_test))
  plot(roc_obj,
       main = sprintf("ROC Curve (Test Set) — AUC = %.3f", auc_r),
       xlab = "False Positive Rate", ylab = "True Positive Rate")
  abline(a = 0, b = 1, lty = 2, col = "grey")
}
```

```{r class-breakdown-r}
# Confusion matrix on test set
tibble(
  Actual    = ifelse(test_data$high_value == 1, "High-Value", "Low-Value"),
  Predicted = ifelse(pred_cls_test == 1,        "Predicted HV", "Predicted LV")
) |>
  count(Actual, Predicted) |>
  pivot_wider(names_from = Predicted, values_from = n, values_fill = 0L) |>
  kable(caption = "Confusion matrix — test set (threshold = 0.50)") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python importance-py}
import shap

# SHAP on test set (LinearExplainer uses train set as background)
explainer = shap.LinearExplainer(log_hv, X_train)
shap_vals = explainer.shap_values(X_test)

mean_abs_shap = np.abs(shap_vals).mean(axis=0)
order = np.argsort(mean_abs_shap)[::-1]

feature_names = ["n_transactions", "total_quantity", "n_unique_items", "days_active", "recency"]
fig, ax = plt.subplots(figsize=(8, 5))
colors = ["#1a9850" if log_hv.coef_[0][i] > 0 else "#d73027" for i in order]
ax.barh(range(len(order)), mean_abs_shap[order], color=colors)
ax.set_yticks(range(len(order)))
ax.set_yticklabels([feature_names[i] for i in order])
ax.set_xlabel("Mean |SHAP value| (test set)")
ax.set_title("Feature Importance — SHAP LinearExplainer (test set)")
plt.tight_layout(); plt.show()
```

```{python shap-waterfall-py}
# SHAP waterfall for one representative high-value prediction
shap_expl = shap.Explanation(
    values=shap_vals,
    base_values=np.full(len(X_test), explainer.expected_value),
    data=X_test,
    feature_names=feature_names
)
hv_idx = int(np.argmax(y_prob))   # most confidently predicted high-value customer
shap.plots.waterfall(shap_expl[hv_idx], max_display=6, show=False)
plt.title(
    f"SHAP Waterfall — Representative High-Value Customer\n"
    f"Predicted probability: {y_prob[hv_idx]:.3f} | True label: {int(y_test.iloc[hv_idx])}"
)
plt.tight_layout(); plt.show()
```

```{python class-roc-py}
from sklearn.metrics import roc_curve, auc

# ROC curve and confusion matrix on TEST set
fpr, tpr, _ = roc_curve(y_test, y_prob)
roc_auc = auc(fpr, tpr)

fig, axes = plt.subplots(1, 2, figsize=(12, 5))

axes[0].plot(fpr, tpr, color="#2c7bb6", linewidth=2, label=f"AUC = {roc_auc:.3f}")
axes[0].plot([0, 1], [0, 1], color="grey", linestyle="--", linewidth=1)
axes[0].set_xlabel("False Positive Rate"); axes[0].set_ylabel("True Positive Rate")
axes[0].set_title("ROC Curve (Test Set)"); axes[0].legend()

cm = confusion_matrix(y_test, y_pred)
sns.heatmap(cm, annot=True, fmt="d", cmap="Blues",
            xticklabels=["Pred LV", "Pred HV"],
            yticklabels=["Actual LV", "Actual HV"], ax=axes[1])
axes[1].set_title("Confusion Matrix (Test Set)")
plt.tight_layout(); plt.show()

print("\nTest Set Classification Report:")
print(classification_report(y_test, y_pred, target_names=["Low-Value", "High-Value"]))
```

:::

**Interpretation:** The confusion matrix is the operational deployment guide for the model. The recall (true positive rate) for the high-value class answers the most commercially important question: of all the customers who are genuinely high-value, what proportion does the model correctly identify? A recall below 60% means the model is missing too many genuine high-value customers to be useful as a priority-setting tool without threshold adjustment. The SHAP values here are computed using `LinearExplainer` — the logistic regression analogue of SHAP — which decomposes each prediction into additive feature contributions. Features with the highest mean absolute SHAP value confirm which purchase behaviours most strongly drive high-value classification and should be communicated to the field sales team as the primary customer-scoring signals.

:::{.callout-note collapse="true"}
## Analytical Notes — Model Evaluation
**What:** AUC, confusion matrix, precision/recall, and SHAP (`LinearExplainer`) feature importance evaluated on the full training dataset.

**Why:** AUC measures discrimination strength across all thresholds; recall specifically measures how many genuine high-value customers are captured at the chosen operating threshold.

**Key assumption:** In-sample evaluation reflects training performance. Cross-validation or a time-based train/test split is recommended before production deployment.

**Management action:** If recall < 60%, lower the classification threshold from 0.5 to 0.4 before deploying scores to the field sales team. Re-evaluate model performance against actual revenue outcomes at 6 months.
:::

---

## 9. Customer Segmentation (Clustering)

**Business justification:** Identifies distinct customer groups (e.g., high-frequency/low-value, low-frequency/high-value, dormant) enabling segment-specific marketing, pricing, and retention strategies.

::: {.panel-tabset}

## R

```{r cluster-r}
# Prepare clustering data — use RFM features (always available)
cluster_data_raw <- df_r |>
  select(total_spend, n_transactions, avg_transaction, days_active, n_unique_items) |>
  drop_na() |>
  mutate(across(everything(), ~ scale(.) |> as.vector()))

# Elbow plot
wss <- map_dbl(1:10, ~ kmeans(cluster_data_raw, centers = ., nstart = 25)$tot.withinss)
ggplot(tibble(k = 1:10, wss = wss), aes(x = k, y = wss)) +
  geom_line() + geom_point(size = 2.5) +
  labs(title = "Elbow Method: Optimal k for Customer Clustering",
       x = "Number of Clusters", y = "Within-Cluster Sum of Squares") +
  theme_minimal()
```

```{r cluster-sil-r}
# Silhouette validation: confirm optimal k
sil_scores_r <- purrr::map_dbl(2:8, function(k) {
  km_k <- kmeans(cluster_data_raw, centers = k, nstart = 25)
  sil_k <- cluster::silhouette(km_k$cluster, dist(cluster_data_raw))
  mean(sil_k[, 3])
})
sil_df_r <- tibble(k = 2:8, silhouette = sil_scores_r)
k_opt_cust <- sil_df_r$k[which.max(sil_df_r$silhouette)]

ggplot(sil_df_r, aes(x = k, y = silhouette)) +
  geom_line() + geom_point(size = 2.5) +
  geom_vline(xintercept = k_opt_cust, linetype = "dashed", colour = "#7b2d8b") +
  annotate("text", x = k_opt_cust + 0.15, y = max(sil_df_r$silhouette),
           label = paste0("Best k = ", k_opt_cust), colour = "#7b2d8b", hjust = 0) +
  labs(title = "Silhouette Score by Number of Clusters",
       x = "Number of Clusters (k)", y = "Mean Silhouette Width") +
  theme_minimal()

cat(sprintf("Best k by silhouette: %d (mean silhouette = %.3f)\n",
            k_opt_cust, max(sil_df_r$silhouette)))
```

```{r cluster-fit-r}
# Fit k-means using k confirmed by both elbow and silhouette methods
km_cust <- kmeans(cluster_data_raw, centers = k_opt_cust, nstart = 25)

# Add clusters to full dataset (only rows used in clustering)
cluster_rows <- which(complete.cases(df_r[, c("total_spend","n_transactions",
                                               "avg_transaction","days_active",
                                               "n_unique_items")]))
df_r$cluster <- NA_integer_
df_r$cluster[cluster_rows] <- km_cust$cluster
df_r_clustered <- df_r |> filter(!is.na(cluster))

# Build profiles, then name clusters by relative mean spend (descending)
seg_labels <- c("Premium Accounts", "Core Buyers", "Occasional Buyers", "Reactivation Targets")

cluster_profiles <- df_r_clustered |>
  group_by(cluster) |>
  summarise(
    n_cust         = n(),
    mean_spend     = mean(total_spend,     na.rm = TRUE),
    mean_freq      = mean(n_transactions,  na.rm = TRUE),
    pct_high_value = mean(high_value,      na.rm = TRUE) * 100,
    .groups = "drop"
  ) |>
  arrange(desc(mean_spend)) |>
  mutate(
    segment = seg_labels[seq_len(n())]
  )

df_r_clustered <- df_r_clustered |>
  left_join(cluster_profiles |> select(cluster, segment), by = "cluster")

cluster_profiles |>
  mutate(across(where(is.numeric), ~ round(., 2))) |>
  kable(caption = "K-Means cluster profiles (segments named by relative mean spend)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r cluster-viz-r}
# 2D viz: Spend vs Frequency, coloured by named segment
df_r_clustered |>
  ggplot(aes(x = total_spend, y = n_transactions, colour = segment)) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_x_continuous(labels = comma) +
  labs(title = "Customer Segments: Total Spend vs Transaction Frequency",
       x = "Total Spend", y = "Number of Transactions",
       colour = "Segment") +
  theme_minimal()
```

## Python

```{python cluster-py}
cluster_py = df_py[["total_spend","n_transactions","avg_transaction","days_active",
                     "n_unique_items"]].dropna()

scaler_cl = StandardScaler()
X_cluster = scaler_cl.fit_transform(cluster_py)

# Elbow
inertias_cl = []
for k in range(1, 11):
    km_k = KMeans(n_clusters=k, n_init=25, random_state=42)
    km_k.fit(X_cluster)
    inertias_cl.append(km_k.inertia_)

fig, ax = plt.subplots(figsize=(8, 4))
ax.plot(range(1, 11), inertias_cl, marker="o")
ax.set_xlabel("k"); ax.set_ylabel("Inertia")
ax.set_title("Elbow Method")
plt.tight_layout(); plt.show()
```

```{python cluster-sil-py}
# Silhouette validation: confirm optimal k
sil_scores_py = []
for k in range(2, 9):
    km_k = KMeans(n_clusters=k, n_init=25, random_state=42)
    labels_k = km_k.fit_predict(X_cluster)
    sil_scores_py.append(silhouette_score(X_cluster, labels_k))

sil_df_py = pd.DataFrame({"k": range(2, 9), "silhouette": sil_scores_py})
k_opt_cust = int(sil_df_py.loc[sil_df_py["silhouette"].idxmax(), "k"])

fig, ax = plt.subplots(figsize=(8, 4))
ax.plot(sil_df_py["k"], sil_df_py["silhouette"], marker="o")
ax.axvline(x=k_opt_cust, color="#7b2d8b", linestyle="--")
ax.set_xlabel("k"); ax.set_ylabel("Mean Silhouette Width")
ax.set_title(f"Silhouette Score by Clusters (best k = {k_opt_cust})")
plt.tight_layout(); plt.show()
print(f"Best k by silhouette: {k_opt_cust} "
      f"(silhouette = {max(sil_scores_py):.3f})")
```

```{python cluster-fit-py}
# Fit k-means using k confirmed by both elbow and silhouette methods
km_cust = KMeans(n_clusters=k_opt_cust, n_init=25, random_state=42)
clusters_cust = km_cust.fit_predict(X_cluster)

cluster_py_temp = cluster_py.copy()
cluster_py_temp["cluster"] = clusters_cust

# Profile clusters
profile_cl = cluster_py_temp.groupby("cluster").agg(
    n_cust    =("total_spend",    "count"),
    mean_spend=("total_spend",    "mean"),
    mean_freq =("n_transactions", "mean"),
    mean_days =("days_active",    "mean")
).round(2).reset_index()

# Name clusters by mean spend (descending rank)
_seg_names = ["Premium Accounts", "Core Buyers",
               "Occasional Buyers", "Reactivation Targets"]
profile_cl_sorted = profile_cl.sort_values("mean_spend", ascending=False).reset_index(drop=True)
profile_cl_sorted["segment"] = [
    _seg_names[i] if i < len(_seg_names) else f"Segment {i+1}"
    for i in range(len(profile_cl_sorted))
]
_seg_map = dict(zip(profile_cl_sorted["cluster"], profile_cl_sorted["segment"]))
cluster_py_temp["segment"] = cluster_py_temp["cluster"].map(_seg_map)
profile_cl["segment"] = profile_cl["cluster"].map(_seg_map)

display(profile_cl[["cluster","segment","n_cust","mean_spend","mean_freq","mean_days"]])
```

```{python cluster-viz-py}
fig, ax = plt.subplots(figsize=(9, 5))
palette = sns.color_palette("tab10", len(_seg_map))
for i, (cid, seg) in enumerate(sorted(_seg_map.items())):
    mask = cluster_py_temp["segment"] == seg
    ax.scatter(cluster_py_temp.loc[mask, "total_spend"],
               cluster_py_temp.loc[mask, "n_transactions"],
               s=40, alpha=0.6, label=seg, color=palette[i])
ax.set_xlabel("Total Spend"); ax.set_ylabel("Transaction Frequency")
ax.set_title("Customer Segments by Spend and Frequency")
ax.legend(title="Segment", bbox_to_anchor=(1.01, 1), loc="upper left")
plt.tight_layout(); plt.show()
```

:::

**Interpretation:** Each cluster should be named and described in terms immediately useful to the commercial team. A cluster of high-frequency, small-basket buyers likely represents active retail accounts with strong product engagement but limited purchase scale — the intervention is basket expansion through range extension and promotional bundling. A cluster of low-frequency, large-basket buyers likely represents periodic wholesale accounts — the intervention is purchase cycle compression through volume discount structures and extended credit terms. Any cluster that combines low spend, low frequency, and short tenure represents customers at risk of attrition and should receive a proactive service check. Cluster centroids on spend and transaction frequency are the primary commercial communication output from this technique.

:::{.callout-note collapse="true"}
## Analytical Notes — Customer Segmentation
**What:** K-Means clustering on standardised RFM features (spend, frequency, quantity, items, tenure). Optimal k selected using combined elbow + silhouette evidence.

**Why:** Elbow identifies the diminishing-return inflection in within-cluster variance; silhouette confirms intra-cluster cohesion relative to inter-cluster separation. Both methods are required for robust k selection.

**Key assumption:** K-Means assumes spherical, similarly-sized clusters in scaled feature space. Non-spherical distributions may warrant hierarchical clustering as a robustness check.

**Management action:** Attach cluster assignment to each customer record in the CRM and use it as the basis for differentiated quarterly promotional planning and territory management.
:::

---

## 10. Dimensionality Reduction (PCA)

**Business justification:** Reduces customer feature space to 2–3 key dimensions, revealing underlying structure in customer heterogeneity and enabling simple rules for customer targeting.

::: {.panel-tabset}

## R

```{r pca-r}
pca_model <- prcomp(cluster_data_raw, center = FALSE, scale. = FALSE)

var_exp <- pca_model$sdev^2 / sum(pca_model$sdev^2)
cumsum_var <- cumsum(var_exp)

tibble(PC = 1:length(var_exp),
       var_pct = var_exp * 100,
       cumsum_pct = cumsum_var * 100) |>
  head(5) |>
  kable(caption = "PCA: Variance explained (%)") |>
  kable_styling(full_width = FALSE)

# Scree plot
ggplot(tibble(PC = seq_along(var_exp), var_pct = var_exp * 100),
       aes(x = PC, y = var_pct)) +
  geom_line() + geom_point(size = 2.5) +
  labs(title = "Scree Plot: PCA Variance Explained",
       x = "Principal Component", y = "Variance (%)") +
  theme_minimal()
```

```{r pca-loadings-r}
# Loadings for PC1 & PC2
loadings_mat <- pca_model$rotation[, 1:2]
rownames(loadings_mat) <- c("Total Spend", "N Transactions", "Avg Trans",
                             "Days Active", "Unique Items")

as.data.frame(loadings_mat) |>
  rownames_to_column("Variable") |>
  kable(digits = 4, caption = "PCA loadings (PC1 & PC2)") |>
  kable_styling(full_width = FALSE)

# Manual biplot with cluster segment overlay
scores_pca <- as.data.frame(pca_model$x[, 1:2])
scores_pca$segment <- df_r_clustered$segment   # attach k-means segment labels
loads_pca  <- as.data.frame(loadings_mat)
loads_pca$var <- rownames(loads_pca)
sf <- max(abs(scores_pca[, 1:2])) / max(abs(loads_pca[, 1:2])) * 0.7

ggplot(scores_pca, aes(x=PC1, y=PC2, colour=segment)) +
  geom_point(alpha=0.4, size=1.5) +
  geom_segment(data=loads_pca, aes(x=0,y=0,xend=PC1*sf,yend=PC2*sf),
    arrow=arrow(length=unit(0.25,"cm")), colour="#e63946", inherit.aes=FALSE) +
  geom_text(data=loads_pca, aes(x=PC1*sf*1.15, y=PC2*sf*1.15, label=var),
    size=3, colour="#e63946", inherit.aes=FALSE) +
  labs(title="PCA Biplot: Feature Contributions & Customer Segments",
       x="PC1", y="PC2", colour="Segment") +
  theme_minimal()
```

## Python

```{python pca-py}
pca_cust = PCA()
pca_cust.fit(X_cluster)

var_exp_pca = pca_cust.explained_variance_ratio_
cumsum_var_pca = np.cumsum(var_exp_pca)

pca_table = pd.DataFrame({
    "PC": range(1, len(var_exp_pca) + 1),
    "Var_%": var_exp_pca * 100,
    "Cumsum_%": cumsum_var_pca * 100
})
display(pca_table.head(5))

n_comp = len(var_exp_pca)
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].plot(range(1, n_comp+1), var_exp_pca * 100, marker="o")
axes[0].set_xlabel("PC"); axes[0].set_ylabel("Variance (%)")
axes[0].set_title("Scree Plot")

axes[1].plot(range(1, n_comp+1), cumsum_var_pca * 100, marker="s")
axes[1].set_xlabel("PC"); axes[1].set_ylabel("Cumsum Variance (%)")
axes[1].set_title("Cumulative Variance")
plt.tight_layout(); plt.show()
```

```{python pca-loadings-py}
loadings_pca = pd.DataFrame(
    pca_cust.components_[:2].T,
    columns=["PC1", "PC2"],
    index=feature_names
)
display(loadings_pca.round(4))
```

```{python pca-cluster-py}
# PCA biplot with cluster segment overlay
X_pca2 = pca_cust.transform(X_cluster)[:, :2]
feature_names_cl = ["total_spend", "n_transactions", "avg_transaction",
                    "days_active", "n_unique_items"]

fig, ax = plt.subplots(figsize=(9, 6))
palette_pca = sns.color_palette("tab10", len(_seg_map))
for i, (cid, seg) in enumerate(sorted(_seg_map.items())):
    mask = (km_cust.labels_ == cid)
    ax.scatter(X_pca2[mask, 0], X_pca2[mask, 1],
               s=30, alpha=0.5, label=seg, color=palette_pca[i])

# Loading arrows
sf_py = (np.abs(X_pca2).max()) / (np.abs(pca_cust.components_[:2]).max()) * 0.5
for j, fname in enumerate(feature_names_cl):
    ax.annotate("", xy=(pca_cust.components_[0, j]*sf_py, pca_cust.components_[1, j]*sf_py),
                xytext=(0, 0),
                arrowprops=dict(arrowstyle="->", color="#e63946", lw=1.5))
    ax.text(pca_cust.components_[0, j]*sf_py*1.15,
            pca_cust.components_[1, j]*sf_py*1.15,
            fname, fontsize=8, color="#e63946")
ax.set_xlabel("PC1"); ax.set_ylabel("PC2")
ax.set_title("PCA Biplot with Customer Segment Overlay")
ax.legend(title="Segment", bbox_to_anchor=(1.01, 1), loc="upper left")
plt.tight_layout(); plt.show()
```

:::

**Interpretation:** The cumulative variance chart shows how many principal components are needed to represent the essential customer information. If the first two components explain more than 50% of total variance, the two-dimensional scatter plot captures the majority of the commercial story. The PC loadings matrix identifies which original features drive each component. PC1 is typically dominated by `total_spend`, `n_transactions`, and `total_quantity` — representing customer scale and commercial activity (‘Revenue Scale’). PC2 is typically driven by `n_unique_items` and `days_active` — representing purchase breadth and customer longevity (‘Engagement Depth’). Overlaying the K-Means cluster assignments on the PC1–PC2 scatter plot visually confirms the cluster separability identified in the previous section: well-separated clusters in PC space validate the K-Means structure and strengthen the commercial case for differentiated segment playbooks.

:::{.callout-note collapse="true"}
## Analytical Notes — PCA
**What:** PCA applied to scaled RFM features (spend, transactions, quantity, items, tenure). First two components summarised in scree plot, loading table, and biplot with cluster overlays.

**Why:** Reduces correlated feature space to interpretable orthogonal dimensions and provides a 2D visualisation confirming cluster separability.

**Key assumption:** PCA is a linear method. Highly skewed or non-linear customer distributions may require t-SNE or UMAP for accurate 2D projection.

**Management action:** Use the PC1–PC2 biplot in the MD’s strategy presentation to communicate the shape of the customer landscape visually and confirm that K-Means clusters are commercially distinct.
:::

---

## 11. Time Series: Monthly Sales Trend, Seasonality & Forecast

**Business justification:** Identifies monthly revenue trends and seasonal patterns (e.g., holiday peaks, summer troughs), informing inventory, promotion, and staffing planning.

::: {.panel-tabset}

## R

```{r ts-r}
# Monthly sales aggregation
monthly_sales <- sales_r |>
  group_by(month_period) |>
  summarise(
    total_amount = sum(amount, na.rm=TRUE),
    n_transactions = n(),
    n_customers = n_distinct(customer_id),
    .groups = "drop"
  ) |>
  arrange(month_period)

# Time series plot
ggplot(monthly_sales, aes(x = month_period, y = total_amount)) +
  geom_line(colour = "#2c7bb6", linewidth = 1) +
  geom_area(fill = "#2c7bb6", alpha = 0.2) +
  scale_y_continuous(labels = comma) +
  labs(title = "Monthly Total Sales Revenue (3 Years)",
       x = "Month", y = "Total Revenue") +
  theme_minimal()
```

```{r ts-decomp-r}
# Prepare time series object
ts_sales <- ts(monthly_sales$total_amount, 
               start = c(year(min(monthly_sales$month_period)), 1),
               frequency = 12)

# Seasonal decomposition
decomp_sales <- decompose(ts_sales, type = "additive")
plot(decomp_sales, xlab = "Time")
```

```{r ts-acf-r}
# ACF and PACF plots — lags capped at 50% of series length
ts_max_lags <- min(24L, floor(length(ts_sales) / 2) - 1L)
par(mfrow = c(1, 2))
acf(ts_sales,  lag.max = ts_max_lags, main = "ACF — Monthly Sales Revenue")
pacf(ts_sales, lag.max = ts_max_lags, main = "PACF — Monthly Sales Revenue")
par(mfrow = c(1, 1))
```

```{r ts-adf-r}
# ADF stationarity test
adf_r <- tseries::adf.test(ts_sales, alternative = "stationary")
cat("=== Augmented Dickey-Fuller Stationarity Test ===\n")
cat(sprintf("  ADF statistic : %.4f\n", as.numeric(adf_r$statistic)))
cat(sprintf("  p-value       : %.4f\n", adf_r$p.value))
cat(sprintf("  Lag order     : %d\n",   as.integer(adf_r$parameter)))
cat(sprintf("  Conclusion    : %s\n",
    if (adf_r$p.value < 0.05)
      "Reject H0 — series is stationary (no unit root)"
    else
      "Fail to reject H0 — non-stationary; differencing may be required"))
```

```{r ts-stats-r}
monthly_sales |>
  mutate(month_label = format(month_period, "%b %Y")) |>
  tail(12) |>
  select(month_label, total_amount, n_transactions, n_customers) |>
  kable(caption = "Last 12 months: revenue, transactions, customer count") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r ts-ets-r}
# ETS forecast with 6-month holdout validation
n_obs_ts  <- length(ts_sales)
n_holdout <- 6
ts_train  <- head(ts_sales, n_obs_ts - n_holdout)
ts_test   <- tail(ts_sales, n_holdout)

ets_fit   <- ets(ts_train)
ets_fcast <- forecast(ets_fit, h = n_holdout)

holdout_rmse <- sqrt(mean((as.numeric(ts_test) -
                           as.numeric(ets_fcast$mean))^2))
holdout_mape <- mean(abs((as.numeric(ts_test) -
                          as.numeric(ets_fcast$mean)) /
                          as.numeric(ts_test))) * 100

cat(sprintf("ETS model: %s\n", ets_fit$method))
cat(sprintf("Holdout RMSE: %s | MAPE: %.1f%%\n",
            formatC(holdout_rmse, format = "f", digits = 0, big.mark = ","),
            holdout_mape))

# 6-month forward forecast from full series
ets_full    <- ets(ts_sales)
ets_forward <- forecast(ets_full, h = 6)

# Build plot data
plot_periods <- seq(min(monthly_sales$month_period),
                    by = "month", length.out = n_obs_ts + 6)
plot_df <- tibble(
  period  = plot_periods,
  revenue = c(as.numeric(ts_sales), as.numeric(ets_forward$mean)),
  lo95    = c(rep(NA_real_, n_obs_ts),
              as.numeric(ets_forward$lower[, "95%"])),
  hi95    = c(rep(NA_real_, n_obs_ts),
              as.numeric(ets_forward$upper[, "95%"])),
  type    = c(rep("Actual", n_obs_ts), rep("Forecast", 6))
)

ggplot(plot_df, aes(x = period)) +
  geom_ribbon(aes(ymin = lo95, ymax = hi95), fill = "#c994c7", alpha = 0.35) +
  geom_line(aes(y = revenue, colour = type, linetype = type), linewidth = 1) +
  scale_colour_manual(values = c("Actual" = "#2c7bb6", "Forecast" = "#7b2d8b")) +
  scale_linetype_manual(values = c("Actual" = "solid", "Forecast" = "dashed")) +
  scale_y_continuous(labels = comma) +
  labs(title = "ETS Forecast: 6-Month Revenue Projection",
       subtitle = sprintf("Holdout RMSE: %s | MAPE: %.1f%%",
                 formatC(holdout_rmse, format = "f", digits = 0, big.mark = ","),
                 holdout_mape),
       x = "Month", y = "Revenue", colour = "", linetype = "") +
  theme_minimal()

# Forecast summary table
tibble(
  Month    = format(seq(max(monthly_sales$month_period) %m+% months(1),
                        by = "month", length.out = 6), "%b %Y"),
  Forecast = comma(round(as.numeric(ets_forward$mean))),
  Lo_95    = comma(round(as.numeric(ets_forward$lower[, "95%"]))),
  Hi_95    = comma(round(as.numeric(ets_forward$upper[, "95%"])))
) |>
  kable(caption = "6-Month Revenue Forecast with 95% Confidence Interval") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python ts-py}
ts_agg = sales_py.groupby("period").agg({
    "amount": "sum",
    "customer_id": "count"
}).reset_index().rename(
    columns={"amount": "total_amount", "customer_id": "n_trans"})
ts_agg = ts_agg.sort_values("period")

fig, ax = plt.subplots(figsize=(11, 4))
ax.plot(ts_agg["period"], ts_agg["total_amount"], color="#2c7bb6", linewidth=1.5)
ax.fill_between(ts_agg["period"], ts_agg["total_amount"], alpha=0.2, color="#2c7bb6")
ax.yaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"${x/1e6:.1f}M"
                                                    if x >= 1e6 else f"${x/1e3:.0f}K"))
ax.set_title("Monthly Total Sales Revenue")
ax.set_xlabel("Month"); ax.set_ylabel("Revenue")
plt.tight_layout(); plt.show()
```

```{python ts-decomp-py}
ts_indexed = ts_agg.set_index("period")["total_amount"]

if len(ts_indexed) >= 2 * 12:
    decomp = seasonal_decompose(ts_indexed, model="additive", period=12)
    
    fig, axes = plt.subplots(4, 1, figsize=(11, 8))
    ts_indexed.plot(ax=axes[0], title="Original", color="#2c7bb6")
    axes[0].set_ylabel("Revenue")
    decomp.trend.plot(ax=axes[1], title="Trend", color="#1a9850")
    axes[1].set_ylabel("Trend")
    decomp.seasonal.plot(ax=axes[2], title="Seasonal", color="#fdae61")
    axes[2].set_ylabel("Seasonal")
    decomp.resid.plot(ax=axes[3], title="Residual", color="#8da0cb")
    axes[3].set_ylabel("Residual")
    plt.tight_layout(); plt.show()
```

```{python ts-acf-py}
# ACF and PACF plots — lags capped at 50% of series length (PACF constraint)
_max_lags = min(24, len(ts_indexed) // 2 - 1)
fig, axes = plt.subplots(1, 2, figsize=(12, 4))
plot_acf( ts_indexed, lags=_max_lags, ax=axes[0], title="ACF — Monthly Sales Revenue")
plot_pacf(ts_indexed, lags=_max_lags, ax=axes[1], title="PACF — Monthly Sales Revenue", method="ywm")
plt.tight_layout(); plt.show()
```

```{python ts-adf-py}
# ADF stationarity test
adf_result = adfuller(ts_indexed.dropna(), autolag="AIC")
print("=== Augmented Dickey-Fuller Stationarity Test ===")
print(f"  ADF statistic : {adf_result[0]:.4f}")
print(f"  p-value       : {adf_result[1]:.4f}")
print(f"  Lags used     : {adf_result[2]}")
print(f"  Conclusion    : {'Reject H0 — series is stationary (no unit root)' if adf_result[1] < 0.05 else 'Fail to reject H0 — non-stationary; differencing may be required'}")
```

```{python ts-ets-py}
# ETS forecast with 6-month holdout validation
n_holdout  = 6
train_ts   = ts_indexed.iloc[:-n_holdout]
test_ts    = ts_indexed.iloc[-n_holdout:]

ets_model  = ExponentialSmoothing(train_ts, trend="add", seasonal="add",
                                   seasonal_periods=12)
ets_fit_py = ets_model.fit(optimized=True)
ets_pred   = ets_fit_py.forecast(n_holdout)

holdout_rmse = np.sqrt(np.mean((test_ts.values - ets_pred.values) ** 2))
holdout_mape = np.mean(np.abs((test_ts.values - ets_pred.values) /
                               test_ts.values)) * 100

print(f"Holdout RMSE: {holdout_rmse:,.0f} | MAPE: {holdout_mape:.1f}%")

# 6-month forward forecast from full series
ets_full     = ExponentialSmoothing(ts_indexed, trend="add", seasonal="add",
                                     seasonal_periods=12)
ets_full_fit = ets_full.fit(optimized=True)
future_idx   = pd.date_range(start=ts_indexed.index[-1], periods=7, freq="MS")[1:]
future_fcast = ets_full_fit.forecast(6)
future_fcast.index = future_idx

fig, ax = plt.subplots(figsize=(11, 4))
ax.plot(ts_indexed.index, ts_indexed.values,
        color="#2c7bb6", linewidth=1.5, label="Actual")
ax.plot(future_idx, future_fcast.values,
        color="#7b2d8b", linewidth=1.5, linestyle="--", label="Forecast")
ax.set_title(
    f"ETS Forecast: 6-Month Revenue Projection\n"
    f"Holdout RMSE: {holdout_rmse:,.0f} | MAPE: {holdout_mape:.1f}%"
)
ax.set_xlabel("Month"); ax.set_ylabel("Revenue")
ax.legend(); plt.tight_layout(); plt.show()

fcast_table = pd.DataFrame({
    "Month":    [d.strftime("%b %Y") for d in future_idx],
    "Forecast": [f"{v:,.0f}" for v in future_fcast.values]
})
display(fcast_table)
```

:::

**Interpretation:** The trend component answers the core planning question: is underlying demand growing, stable, or in structural decline? A positive trend slope across 2023–2025 validates continued investment in trade marketing and distribution expansion. A flat or declining trend would signal a need to review pricing strategy or product range competitiveness before the next annual budget cycle. The seasonal component identifies the highest-demand months — in the Nigerian food distribution context, festive periods (October–December) and school reopening cycles are typical seasonal peaks, which should drive production scheduling, pre-positioning of stock, and advance promotional planning. The 6-month ETS forecast provides the forward revenue baseline, replacing static percentage-uplift assumptions in budget planning. The RMSE and MAPE on the holdout period confirm how tightly the model tracks actual performance, anchoring confidence in the forecast range.

:::{.callout-note collapse="true"}
## Analytical Notes — Time Series
**What:** Classical seasonal decomposition (trend, seasonal, residual) and ETS forecast fitted on 36 months of monthly revenue; last 6 months held out for validation; 6-month forward forecast generated from full series.

**Why:** ETS automatically selects the optimal error, trend, and seasonal specification; RMSE/MAPE on the holdout period provides an honest accuracy estimate before forward projections are used in planning.

**Key assumption:** Future seasonal pattern resembles past pattern. Structural disruptions — new product lines, distribution changes, major macroeconomic shocks — not captured in historical series will reduce forecast accuracy.

**Management action:** Use the 6-month forecast midpoint as the revenue baseline in the next budget planning round; treat the 95% CI lower and upper bounds as conservative and optimistic scenario floors for financial planning.
:::

---

## 12. Integrated Findings

Five analytical techniques converge on a customer intelligence architecture that transforms three years of sales data into an actionable commercial plan for the food company’s distribution operations. The classification model established which customers are on a high-value trajectory using behavioural RFM predictors — frequency, quantity, product breadth, tenure, and recency — providing the prioritisation engine for field sales investment. Model evaluation validated the model’s operational reliability via AUC, confusion matrix, and SHAP feature contributions, translating statistical outputs into feature-level guidance for the sales team. Customer segmentation revealed distinct commercial profiles within the customer base, enabling differentiated playbooks rather than uniform distribution strategy. PCA confirmed the cluster structure in a reduced feature space and identified the two dominant dimensions of customer variation — commercial scale and engagement depth — providing the managing director with a visually communicable customer landscape map. The time series analysis placed customer intelligence in its temporal context: the trend trajectory, seasonal pattern, and 6-month ETS forecast provide the revenue planning foundation that anchors all commercial interventions in a realistic forward outlook.

**Recommendation:** Deploy a three-tier customer engagement model anchored in the classification and clustering results. The top-tier (predicted high-value customers in the highest-spend cluster) receives dedicated key account management and priority product access; the mid-tier receives quarterly promotional programmes calibrated to the seasonal peaks identified in the time series; the low-tier receives cost-effective digital interventions (SMS promotions, digital catalogues) with reactivation triggers based on recency score. Review model performance against actual revenue outcomes at the 6-month mark and recalibrate cluster assignments and classification thresholds accordingly.

---

## 13. Limitations & Further Work

- **Survey representativeness:** The 190 survey responses cannot be matched to the 8,017 sales customers (different instruments, no shared identifier). Survey data is used descriptively only; predictive models rely exclusively on sales-derived RFM features.
- **Temporal dynamics:** Clustering is static; in reality, customers move between segments over time (e.g., frequent buyers becoming dormant). A transition matrix or hidden Markov model would capture migration dynamics.
- **Revenue vs profit:** The analyses focus on revenue; profitability by customer segment may differ substantially due to cost-of-goods, acquisition cost, or logistics variation.
- **Forecast horizon:** 36 months is at the lower bound recommended for seasonal ETS modelling. Forward forecast confidence intervals will be wide; interpret directional trend and seasonal patterns rather than relying on precise point forecasts.
- **Future work:** Build a customer lifetime value (CLV) model integrating acquisition cost, retention probability, and repeat purchase rate to score customers on expected profitability over 12–36 months.

---

## Executive Dashboard

::: {.panel-tabset}

## R

```{r dashboard-r}
#| echo: false
tibble(
  KPI = c(
    "Total customers analysed",
    "High-value customers (%)",
    "Classification AUC",
    "Optimal clusters (k)",
    "ETS forecast RMSE (holdout)",
    "ETS forecast MAPE (holdout)"
  ),
  Value = c(
    formatC(nrow(df_r), format = "d", big.mark = ","),
    sprintf("%.1f%%", mean(df_r$high_value, na.rm = TRUE) * 100),
    tryCatch(sprintf("%.3f", auc_r), error = function(e) "N/A"),
    as.character(k_opt_cust),
    tryCatch(formatC(holdout_rmse, format = "f", digits = 0, big.mark = ","), error = function(e) "N/A"),
    tryCatch(sprintf("%.1f%%", holdout_mape), error = function(e) "N/A")
  )
) |>
  kable(caption = "Executive Analytics Dashboard — key outputs across all five techniques") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python dashboard-py}
#| echo: false
from sklearn.metrics import roc_auc_score as _auc_score
_auc_val = f"{_auc_score(y_test, y_prob):.3f}"
_dashboard = {
    "Total customers analysed": f"{len(df_py):,}",
    "High-value customers (%)": f"{df_py['high_value'].mean()*100:.1f}%",
    "Classification AUC": _auc_val,
    "Optimal clusters (k)": str(k_opt_cust),
    "ETS forecast RMSE (holdout)": f"{holdout_rmse:,.0f}",
    "ETS forecast MAPE (holdout)": f"{holdout_mape:.1f}%"
}
display(pd.DataFrame(list(_dashboard.items()), columns=["KPI", "Value"]))
```

:::

---

## Reproducibility & Package Citations

::: {.panel-tabset}

## R

```{r repro-r}
#| echo: false
r_pkgs_list <- c("readr", "dplyr", "tidyr", "tibble", "purrr", "forcats",
             "ggplot2", "lubridate", "scales", "broom", "knitr",
             "cluster", "forecast", "zoo", "pROC", "kableExtra")
pkg_info_r <- purrr::map_dfr(r_pkgs_list, function(p) {
  v <- tryCatch(as.character(packageVersion(p)), error = function(e) "not installed")
  tibble(Package = p, Version = v)
})
pkg_info_r |>
  kable(caption = paste0("R ", R.version$major, ".", R.version$minor,
                          " — installed package versions")) |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python repro-py}
#| echo: false
import importlib.metadata as _imeta
_py_pkgs = ["pandas", "numpy", "matplotlib", "seaborn", "scipy",
            "scikit-learn", "statsmodels", "shap"]
_pkg_rows = []
for _p in _py_pkgs:
    try:
        _pkg_rows.append({"Package": _p, "Version": _imeta.version(_p)})
    except Exception:
        _pkg_rows.append({"Package": _p, "Version": "not found"})
display(pd.DataFrame(_pkg_rows))
```

:::

---

## GitHub Repository

The code and anonymised data for this analysis are available at:

**Repository:** `https://github.com/oluchi-frances/customer-analytics-capstone` *

```
customer-analytics-capstone/
├── oluchi_capstone.qmd          # Main analysis document
├── oluchi_report.css            # Report stylesheet
├── requirements.txt             # Python dependencies
├── data/
│   ├── sales_data_3yrs.csv      # 3-year sales transactions (8,017 rows)
│   └── customer_survey_data.csv # 190 survey responses (descriptive only)
└── README.md                    # Project overview and render instructions
```

---

## References

- Adi, B. (2026). *AI-powered business analytics: A practical textbook for data-driven decision making — from data fundamentals to machine learning in Python and R*. Lagos Business School / markanalytics.online. https://markanalytics.online
- R Core Team. (2024). *R: A language and environment for statistical computing*. R Foundation for Statistical Computing.
- Wickham, H., & Grolemund, G. (2017). *R for Data Science*. O'Reilly Media.
- McKinney, W. (2022). *Python for Data Analysis* (3rd ed.). O'Reilly Media.
- Lundberg, S. M., & Lee, S. I. (2017). A unified approach to interpreting model predictions. *Advances in Neural Information Processing Systems*, 30.
- Pedersen, T. L. (2024). *patchwork: The Composer of Plots* (R package version 1.2.0). CRAN.
- Xie, Y. (2024). *knitr: A General-Purpose Package for Dynamic Report Generation in R* (R package). CRAN.

---

## Appendix: AI Usage Statement

GitHub Copilot and ChatGPT were used as drafting and code-structuring tools during the preparation of this capstone. The analytical design — including the leakage-safe feature selection (excluding `total_spend` and `avg_transaction` from classification predictors), the decision to treat survey data descriptively rather than joining it to the sales record set, the ETS holdout evaluation approach, and the silhouette-confirmed cluster selection — was independently developed and validated by the author using the company’s own sales and survey data and the outputs generated from the models. Responsibility for the analysis, interpretations, and commercial recommendations rests entirely with the author.
