---
title: "Customer Analytics: Segmentation, Lifetime Value Prediction, and Personalisation Strategy"
author: "Oluchi Frances"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
pkgs <- c("tidyverse", "lubridate", "scales", "ggcorrplot",
          "patchwork", "car", "broom", "knitr", "kableExtra", "janitor",
          "cluster", "factoextra")
invisible(lapply(pkgs, library, character.only = TRUE))
knitr::opts_chunk$set(fig.width = 9, fig.height = 5, dpi = 150)
```

```{python setup-py, include=FALSE}
import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
import matplotlib.ticker as mticker
import seaborn as sns
from scipy.stats import chi2_contingency, pointbiserialr
from sklearn.preprocessing import StandardScaler
from sklearn.cluster import KMeans
from sklearn.decomposition import PCA
from sklearn.linear_model import LogisticRegression
from sklearn.metrics import (classification_report, confusion_matrix, roc_auc_score,
                             roc_curve, auc, accuracy_score)
import statsmodels.api as sm
from statsmodels.tsa.seasonal import seasonal_decompose
import warnings
warnings.filterwarnings("ignore")
from IPython.display import display
sns.set_theme(style="whitegrid", palette="muted")
plt.rcParams["figure.dpi"] = 150
```

```{r load-r, include=FALSE}
# Load sales data
sales_r <- read_csv("data/sales_data_3yrs.csv",
                    col_types = cols(.default = col_character())) |>
  clean_names() |>
  mutate(
    customer_id = as.character(customer_id),
    item_id     = as.character(item_id),
    qty         = as.numeric(qty),
    amount      = as.numeric(gsub("[^0-9.]", "", amount)),
    period      = my(period),
    month_period = floor_date(period, "month")
  ) |>
  filter(!is.na(amount))

# Load survey data
survey_r <- read_csv("data/customer_survey_data.csv",
                     col_types = cols(.default = col_character())) |>
  clean_names() |>
  mutate(
    timestamp = mdy_hms(timestamp, quiet=TRUE),
    customer_id = as.character(row_number())  # Map survey row → pseudo-customer ID
  )

# Aggregate sales by customer (RFM-style features)
customer_rfm <- sales_r |>
  group_by(customer_id) |>
  summarise(
    n_transactions = n(),
    total_spend    = sum(amount, na.rm=TRUE),
    avg_transaction = mean(amount, na.rm=TRUE),
    total_quantity = sum(qty, na.rm=TRUE),
    n_unique_items = n_distinct(item_id),
    first_purchase = min(month_period, na.rm=TRUE),
    last_purchase  = max(month_period, na.rm=TRUE),
    days_active    = as.numeric(difftime(max(month_period, na.rm=TRUE),
                                         min(month_period, na.rm=TRUE),
                                         units = "days")),
    .groups = "drop"
  ) |>
  filter(!is.na(total_spend))

# Merge with survey data (best-effort join on row order)
survey_features <- survey_r |>
  select(customer_id, age_group, income, product_familiarity,
         purchase_frequency) |>
  mutate(
    customer_id = as.character(customer_id),
    # Encode categorical as ordinal
    age_group_num = case_when(
      age_group == "18-25" ~ 1,
      age_group == "26-35" ~ 2,
      age_group == "36-50" ~ 3,
      age_group == "50+" ~ 4,
      TRUE ~ NA_real_
    ),
    income_num = case_when(
      income == "Low" ~ 1,
      income == "Medium" ~ 2,
      income == "High" ~ 3,
      TRUE ~ NA_real_
    ),
    familiarity_num = case_when(
      product_familiarity == "Low" ~ 1,
      product_familiarity == "Medium" ~ 2,
      product_familiarity == "High" ~ 3,
      TRUE ~ NA_real_
    ),
    freq_num = case_when(
      purchase_frequency == "Occasional" ~ 1,
      purchase_frequency == "Regular" ~ 2,
      purchase_frequency == "Frequent" ~ 3,
      TRUE ~ NA_real_
    )
  )

# Create integrated dataset
df_r <- customer_rfm |>
  left_join(survey_features, by = "customer_id") |>
  mutate(
    high_value = if_else(total_spend >= median(total_spend, na.rm=TRUE), 1, 0)
  )
```

```{python load-py, include=FALSE}
# Sales data
sales_py = pd.read_csv("data/sales_data_3yrs.csv", dtype=str)
sales_py.columns = sales_py.columns.str.strip()
sales_py.columns = (sales_py.columns.str.lower()
                    .str.replace(r"[^\w]", "_", regex=True)
                    .str.replace(r"_+", "_", regex=True)
                    .str.strip("_"))

sales_py["amount"] = pd.to_numeric(
    sales_py["amount"].str.replace(r"[^\d.\-]", "", regex=True), errors="coerce")
sales_py["qty"] = pd.to_numeric(sales_py["qty"], errors="coerce")
sales_py["period"] = pd.to_datetime(sales_py["period"].str.strip(), 
                                     format="%b-%y", errors="coerce")
sales_py = sales_py.dropna(subset=["amount"])

# Customer RFM features
rfm_py = sales_py.groupby("customer_id").agg(
    n_transactions=("customer_id", "count"),
    total_spend=("amount", "sum"),
    avg_transaction=("amount", "mean"),
    total_quantity=("qty", "sum"),
    n_unique_items=("item_id", "nunique"),
    first_purchase=("period", "min"),
    last_purchase=("period", "max")
).reset_index()

rfm_py["days_active"] = (rfm_py["last_purchase"] - 
                         rfm_py["first_purchase"]).dt.days

# Survey data
survey_py = pd.read_csv("data/customer_survey_data.csv", dtype=str)
survey_py.columns = survey_py.columns.str.strip()
survey_py.columns = (survey_py.columns.str.lower()
                     .str.replace(r"[^\w]", "_", regex=True)
                     .str.replace(r"_+", "_", regex=True)
                     .str.strip("_"))

survey_py["customer_id"] = range(1, len(survey_py) + 1)

# Encode categoricals
age_map = {"18-25": 1, "26-35": 2, "36-50": 3, "50+": 4}
income_map = {"Low": 1, "Medium": 2, "High": 3}
fam_map = {"Low": 1, "Medium": 2, "High": 3}
freq_map = {"Occasional": 1, "Regular": 2, "Frequent": 3}

survey_py["age_group_num"] = survey_py.get("age_group", "").map(age_map)
survey_py["income_num"] = survey_py.get("income", "").map(income_map)
survey_py["familiarity_num"] = survey_py.get("product_familiarity", "").map(fam_map)
survey_py["freq_num"] = survey_py.get("purchase_frequency", "").map(freq_map)

# Merge
df_py = rfm_py.merge(survey_py[["customer_id","age_group_num","income_num",
                                  "familiarity_num","freq_num"]],
                     on="customer_id", how="left")

df_py["high_value"] = (df_py["total_spend"] >= 
                       df_py["total_spend"].median()).astype(int)
```

---

## 1. Executive Summary

> **Instructions:** Complete after your analysis. 150–200 words.

**Business problem:** [Food company name]'s sales are driven by a diverse customer base spanning different age groups, income levels, and purchase behaviours. Understanding which customer segments generate the highest lifetime value and what characteristics predict high-value status enables targeted acquisition, pricing, and retention strategies.

**Data collected:** 8,017 sales transactions covering three years (January 2023–December 2025) and 190 customer survey responses capturing demographics, product familiarity, and purchase frequency.

**Key findings:**

1. The logistic regression model for predicting high-value customers (above median spend) achieves [AUC/accuracy metrics].
2. K-means clustering identifies [N] distinct customer segments; the "premium loyal" segment (Cluster X) accounts for [Y%] of total revenue despite representing [Z%] of customer base.
3. PCA reveals that the first two components capture [X%] of variance, driven primarily by [spending scale / income level / purchase frequency].
4. Monthly sales show [upward/seasonal] pattern with [seasonality strength]% seasonal variation.

**Recommendation:** [Specific customer segmentation, targeting, or retention action.]

---

## 2. Professional Disclosure

> **Instructions:** Paste your completed Section 2 from `docs/section_a_30_marks_template.md` here.

---

## 3. Data Collection & Sampling

> **Instructions:** Paste your completed Section 3 from `docs/section_a_30_marks_template.md` here.

---

## 4. Data Description

::: {.panel-tabset}

## R

```{r desc-r}
cat("Sales transactions:", nrow(sales_r), "| Survey responses:", nrow(survey_r), "\n")
cat("Period:", format(min(sales_r$period, na.rm=TRUE), "%b %Y"),
    "to", format(max(sales_r$period, na.rm=TRUE), "%b %Y"), "\n\n")

df_r |>
  select(total_spend, n_transactions, avg_transaction, total_quantity,
         days_active, age_group_num, income_num) |>
  summary() |>
  kable(caption = "Customer-level and survey variable summaries") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

cat("\nHigh-Value Customer Distribution:\n")
df_r |> count(high_value) |>
  mutate(high_value = if_else(high_value == 1, "High-Value", "Low-Value"),
         pct = n / sum(n) * 100) |>
  kable(caption = "Customers above/below median spend") |>
  kable_styling(full_width = FALSE)
```

```{r desc-cats-r}
df_r |> count(age_group, sort=TRUE) |>
  kable(caption = "Survey respondents by age group") |>
  kable_styling(full_width = FALSE)

df_r |> count(income, sort=TRUE) |>
  kable(caption = "Survey respondents by income level") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python desc-py}
print(f"Sales transactions: {len(sales_py)} | Survey responses: {len(survey_py)}")
print(f"Sales period: {sales_py['period'].min().strftime('%b %Y')} "
      f"to {sales_py['period'].max().strftime('%b %Y')}\n")

display(df_py[["total_spend","n_transactions","avg_transaction","total_quantity",
               "days_active","age_group_num","income_num"]].describe().round(2))

print("\nHigh-Value distribution:")
hv_counts = df_py["high_value"].value_counts().sort_index()
print(f"Low-Value: {hv_counts.get(0, 0)} | High-Value: {hv_counts.get(1, 0)}")
```

```{python desc-cats-py}
print("Age group (survey):")
display(survey_py["age_group"].value_counts().reset_index(name="count"))
print("\nIncome level (survey):")
display(survey_py["income"].value_counts().reset_index(name="count"))
```

:::

---

## 5. Technique 1 — Classification: Predicting High-Value Customers

**Business justification:** Builds a predictive model for customer lifetime value potential (above/below median spend), enabling proactive targeting of high-potential customers for premium offerings and loyalty programmes.

::: {.panel-tabset}

## R

```{r class-r}
# Prepare data
class_data <- df_r |>
  select(high_value, total_spend, n_transactions, avg_transaction,
         age_group_num, income_num, familiarity_num, freq_num) |>
  drop_na()

# Logistic regression
glm_hv <- glm(high_value ~ total_spend + age_group_num + income_num +
              familiarity_num + freq_num,
              family = binomial, data = class_data)
cat("=== Logistic Regression: High-Value Customer Prediction ===\n")
summary(glm_hv)
```

```{r class-table-r}
tidy(glm_hv, exponentiate = TRUE, conf.int = TRUE) |>
  mutate(across(where(is.numeric), ~ round(., 4))) |>
  kable(caption = "Logistic regression coefficients (odds ratios)") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

# Performance
pred_prob <- predict(glm_hv, type = "response")
auc_val <- pROC::auc(class_data$high_value, pred_prob)
cat(sprintf("\nROC-AUC: %.4f\n", auc_val))

pred_class <- if_else(pred_prob >= 0.5, 1, 0)
acc <- mean(pred_class == class_data$high_value)
cat(sprintf("Accuracy (threshold = 0.5): %.4f\n", acc))
```

## Python

```{python class-py}
from sklearn.metrics import roc_auc_score, accuracy_score, confusion_matrix

class_py = df_py[["high_value","total_spend","n_transactions","avg_transaction",
                   "age_group_num","income_num","familiarity_num","freq_num"]].dropna()

X_c = class_py[["total_spend","n_transactions","avg_transaction",
                 "age_group_num","income_num","familiarity_num","freq_num"]]
y_c = class_py["high_value"]

# Standardise
scaler_c = StandardScaler()
X_c_scaled = scaler_c.fit_transform(X_c)

# Logistic regression
log_hv = LogisticRegression(random_state=42, max_iter=1000)
log_hv.fit(X_c_scaled, y_c)

print("=== Logistic Regression: High-Value Customer ===")
print(f"Coefficients: {np.round(log_hv.coef_[0], 4)}")
print(f"Intercept: {log_hv.intercept_[0]:.4f}")

y_prob = log_hv.predict_proba(X_c_scaled)[:, 1]
y_pred = log_hv.predict(X_c_scaled)

auc_c = roc_auc_score(y_c, y_prob)
acc_c = accuracy_score(y_c, y_pred)
print(f"\nROC-AUC: {auc_c:.4f}")
print(f"Accuracy: {acc_c:.4f}")
```

:::

> **Interpretation:** Which features most strongly increase odds of being high-value (positive coefficient)? Is the AUC > 0.7? Translate this into a business rule: "Customers with [feature X above threshold] have [Z times] higher odds of being high-value."

---

## 6. Technique 2 — Model Evaluation & Feature Importance

**Business justification:** Visualises which customer characteristics have the greatest impact on high-value prediction, enabling marketing teams to design targeted acquisition profiles and messaging.

::: {.panel-tabset}

## R

```{r importance-r}
# Coefficient magnitude as importance
importance_df <- tidy(glm_hv) |>
  filter(term != "(Intercept)") |>
  mutate(abs_estimate = abs(estimate)) |>
  arrange(desc(abs_estimate))

ggplot(importance_df, aes(x = reorder(term, abs_estimate), y = estimate,
                           fill = sign(estimate) > 0)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("#d73027", "#1a9850")) +
  coord_flip() +
  labs(title = "Feature Importance: Drivers of High-Value Customer Status",
       subtitle = "Red = decreases odds | Green = increases odds",
       x = "Feature", y = "Coefficient (log-odds)") +
  theme_minimal()

# ROC curve
roc_obj <- pROC::roc(class_data$high_value, pred_prob)
plot(roc_obj, main = "ROC Curve: High-Value Customer Prediction",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
abline(a = 0, b = 1, lty = 2, col = "grey")
```

```{r class-breakdown-r}
# Classification table by actual spend quartile
df_r_pred <- df_r |> drop_na(contains("_num")) |>
  mutate(pred_prob = predict(glm_hv, newdata = ., type = "response"),
         pred_class = if_else(pred_prob >= 0.5, "Predicted HV", "Predicted LV"))

df_r_pred |>
  group_by(high_value, pred_class) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = pred_class, values_from = n, values_fill = 0) |>
  kable(caption = "Classification matrix") |>
  kable_styling(full_width = FALSE)
```

## Python

```{python importance-py}
import shap

# SHAP values
explainer = shap.LinearExplainer(log_hv, X_c_scaled)
shap_vals = explainer.shap_values(X_c_scaled)

mean_abs_shap = np.abs(shap_vals).mean(axis=0)
order = np.argsort(mean_abs_shap)[::-1]

fig, ax = plt.subplots(figsize=(8, 5))
colors = ["#1a9850" if log_hv.coef_[0][i] > 0 else "#d73027" for i in order]
ax.barh(range(len(order)), mean_abs_shap[order], color=colors)
feature_names = ["total_spend", "n_trans", "avg_trans", "age_num", "income_num", "fam_num", "freq_num"]
ax.set_yticks(range(len(order)))
ax.set_yticklabels([feature_names[i] for i in order])
ax.set_xlabel("Mean |SHAP value|")
ax.set_title("Feature Importance (SHAP values)")
plt.tight_layout(); plt.show()
```

```{python class-roc-py}
from sklearn.metrics import roc_curve, auc

fpr, tpr, _ = roc_curve(y_c, y_prob)
roc_auc = auc(fpr, tpr)

fig, ax = plt.subplots(figsize=(7, 6))
ax.plot(fpr, tpr, color="#2c7bb6", linewidth=2, label=f"AUC = {roc_auc:.3f}")
ax.plot([0, 1], [0, 1], color="grey", linestyle="--", linewidth=1)
ax.set_xlabel("False Positive Rate"); ax.set_ylabel("True Positive Rate")
ax.set_title("ROC Curve: High-Value Customer Prediction")
ax.legend(); plt.tight_layout(); plt.show()
```

:::

> **Interpretation:** Which feature has the highest SHAP impact? Translate the ROC curve into plain language: "The model correctly identifies [X%] of high-value customers while falsely flagging [Y%] of low-value customers as high-value."

---

## 7. Technique 3 — Clustering: Customer Segmentation

**Business justification:** Identifies distinct customer groups (e.g., high-frequency/low-value, low-frequency/high-value, dormant) enabling segment-specific marketing, pricing, and retention strategies.

::: {.panel-tabset}

## R

```{r cluster-r}
# Prepare clustering data
cluster_data_raw <- df_r |>
  select(total_spend, n_transactions, avg_transaction, days_active,
         age_group_num, income_num, familiarity_num, freq_num) |>
  drop_na() |>
  mutate(across(everything(), ~ scale(.) |> as.vector()))

# Elbow plot
wss <- map_dbl(1:10, ~ kmeans(cluster_data_raw, centers = ., nstart = 25)$tot.withinss)
ggplot(tibble(k = 1:10, wss = wss), aes(x = k, y = wss)) +
  geom_line() + geom_point(size = 2.5) +
  labs(title = "Elbow Method: Optimal k for Customer Clustering",
       x = "Number of Clusters", y = "Within-Cluster Sum of Squares") +
  theme_minimal()
```

```{r cluster-fit-r}
# Fit k-means (adjust k based on elbow)
k_opt_cust <- 4
km_cust <- kmeans(cluster_data_raw, centers = k_opt_cust, nstart = 25)

# Add clusters to full dataset
df_r_clustered <- df_r |> drop_na(contains("_num")) |>
  mutate(cluster = km_cust$cluster)

# Cluster profiles
df_r_clustered |>
  group_by(cluster) |>
  summarise(
    n_cust = n(),
    mean_spend = mean(total_spend, na.rm=TRUE),
    mean_freq = mean(n_transactions, na.rm=TRUE),
    pct_high_value = mean(high_value, na.rm=TRUE) * 100,
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), ~round(., 2))) |>
  kable(caption = "K-Means cluster profiles") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

```{r cluster-viz-r}
# 2D viz: Spend vs Frequency by cluster
df_r_clustered |>
  ggplot(aes(x = total_spend, y = n_transactions, colour = factor(cluster))) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_x_continuous(labels = comma) +
  labs(title = "Customer Clusters: Total Spend vs Transaction Frequency",
       x = "Total Spend", y = "Number of Transactions",
       colour = "Cluster") +
  theme_minimal()
```

## Python

```{python cluster-py}
cluster_py = df_py[["total_spend","n_transactions","avg_transaction","days_active",
                     "age_group_num","income_num","familiarity_num","freq_num"]].dropna()

scaler_cl = StandardScaler()
X_cluster = scaler_cl.fit_transform(cluster_py)

# Elbow
inertias_cl = []
for k in range(1, 11):
    km_k = KMeans(n_clusters=k, n_init=25, random_state=42)
    km_k.fit(X_cluster)
    inertias_cl.append(km_k.inertia_)

fig, ax = plt.subplots(figsize=(8, 4))
ax.plot(range(1, 11), inertias_cl, marker="o")
ax.set_xlabel("k"); ax.set_ylabel("Inertia")
ax.set_title("Elbow Method")
plt.tight_layout(); plt.show()
```

```{python cluster-fit-py}
k_opt_cust = 4
km_cust = KMeans(n_clusters=k_opt_cust, n_init=25, random_state=42)
clusters_cust = km_cust.fit_predict(X_cluster)

cluster_py_temp = cluster_py.copy()
cluster_py_temp["cluster"] = clusters_cust

summary_cl = cluster_py_temp.groupby("cluster").agg({
    "total_spend": ["count", "mean"],
    "n_transactions": "mean",
    "age_group_num": "mean"
}).round(2)
display(summary_cl)
```

```{python cluster-viz-py}
fig, ax = plt.subplots(figsize=(9, 5))
scatter = ax.scatter(cluster_py["total_spend"], cluster_py["n_transactions"],
                     c=clusters_cust, s=40, alpha=0.6, cmap="viridis")
ax.set_xlabel("Total Spend"); ax.set_ylabel("Transaction Frequency")
ax.set_title("Customer Clusters")
plt.colorbar(scatter, label="Cluster")
plt.tight_layout(); plt.show()
```

:::

> **Interpretation:** Describe each cluster in business terms. Example: "Cluster 1 (n=245): Low-spend, high-frequency shoppers — 'value hunters' seeking regular small purchases. Cluster 2 (n=89): High-spend, occasional buyers — 'premium occasional' segment, ripe for loyalty rewards."

---

## 8. Technique 4 — Principal Component Analysis (PCA)

**Business justification:** Reduces customer feature space to 2–3 key dimensions, revealing underlying structure in customer heterogeneity and enabling simple rules for customer targeting.

::: {.panel-tabset}

## R

```{r pca-r}
pca_model <- prcomp(cluster_data_raw, center = FALSE, scale. = FALSE)

var_exp <- pca_model$sdev^2 / sum(pca_model$sdev^2)
cumsum_var <- cumsum(var_exp)

tibble(PC = 1:length(var_exp),
       var_pct = var_exp * 100,
       cumsum_pct = cumsum_var * 100) |>
  head(5) |>
  kable(caption = "PCA: Variance explained (%)") |>
  kable_styling(full_width = FALSE)

# Scree plot
ggplot(tibble(PC = 1:8, var_pct = var_exp[1:8] * 100),
       aes(x = PC, y = var_pct)) +
  geom_line() + geom_point(size = 2.5) +
  labs(title = "Scree Plot: PCA Variance Explained",
       x = "Principal Component", y = "Variance (%)") +
  theme_minimal()
```

```{r pca-loadings-r}
# Loadings for PC1 & PC2
loadings_mat <- pca_model$rotation[, 1:2]
rownames(loadings_mat) <- c("Total Spend", "N Transactions", "Avg Trans",
                             "Days Active", "Age", "Income", "Familiarity", "Frequency")

as.data.frame(loadings_mat) |>
  rownames_to_column("Variable") |>
  kable(digits = 4, caption = "PCA loadings (PC1 & PC2)") |>
  kable_styling(full_width = FALSE)

# Biplot
autoplot(pca_model, loadings = TRUE, loadings.label = TRUE,
         loadings.label.size = 3, scale = 0) +
  labs(title = "PCA Biplot: Feature Contributions to PC1 & PC2") +
  theme_minimal()
```

## Python

```{python pca-py}
pca_cust = PCA()
pca_cust.fit(X_cluster)

var_exp_pca = pca_cust.explained_variance_ratio_
cumsum_var_pca = np.cumsum(var_exp_pca)

pca_table = pd.DataFrame({
    "PC": range(1, len(var_exp_pca) + 1),
    "Var_%": var_exp_pca * 100,
    "Cumsum_%": cumsum_var_pca * 100
})
display(pca_table.head(5))

fig, axes = plt.subplots(1, 2, figsize=(12, 4))
axes[0].plot(range(1, 9), var_exp_pca[:8] * 100, marker="o")
axes[0].set_xlabel("PC"); axes[0].set_ylabel("Variance (%)")
axes[0].set_title("Scree Plot")

axes[1].plot(range(1, 9), cumsum_var_pca[:8] * 100, marker="s")
axes[1].set_xlabel("PC"); axes[1].set_ylabel("Cumsum Variance (%)")
axes[1].set_title("Cumulative Variance")
plt.tight_layout(); plt.show()
```

```{python pca-loadings-py}
loadings_pca = pd.DataFrame(
    pca_cust.components_[:2].T,
    columns=["PC1", "PC2"],
    index=feature_names
)
display(loadings_pca.round(4))
```

:::

> **Interpretation:** PC1 typically captures scale (spending magnitude); PC2 often captures frequency or engagement. Example: "PC1 explains 38% of variance and is driven by total spend and transaction count — representing 'customer scale'. PC2 explains 22% and is dominated by product familiarity — representing 'market engagement'."

---

## 9. Technique 5 — Time Series: Monthly Sales Trend & Seasonality

**Business justification:** Identifies monthly revenue trends and seasonal patterns (e.g., holiday peaks, summer troughs), informing inventory, promotion, and staffing planning.

::: {.panel-tabset}

## R

```{r ts-r}
# Monthly sales aggregation
monthly_sales <- sales_r |>
  group_by(month_period) |>
  summarise(
    total_amount = sum(amount, na.rm=TRUE),
    n_transactions = n(),
    n_customers = n_distinct(customer_id),
    .groups = "drop"
  ) |>
  arrange(month_period)

# Time series plot
ggplot(monthly_sales, aes(x = month_period, y = total_amount)) +
  geom_line(colour = "#2c7bb6", linewidth = 1) +
  geom_area(fill = "#2c7bb6", alpha = 0.2) +
  scale_y_continuous(labels = comma) +
  labs(title = "Monthly Total Sales Revenue (3 Years)",
       x = "Month", y = "Total Revenue") +
  theme_minimal()
```

```{r ts-decomp-r}
# Prepare time series object
ts_sales <- ts(monthly_sales$total_amount, 
               start = c(year(min(monthly_sales$month_period)), 1),
               frequency = 12)

# Seasonal decomposition
decomp_sales <- decompose(ts_sales, type = "additive")
plot(decomp_sales, xlab = "Time")
```

```{r ts-stats-r}
monthly_sales |>
  mutate(month_label = format(month_period, "%b %Y")) |>
  tail(12) |>
  select(month_label, total_amount, n_transactions, n_customers) |>
  kable(caption = "Last 12 months: revenue, transactions, customer count") |>
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Python

```{python ts-py}
ts_agg = sales_py.groupby("period").agg({
    "amount": "sum",
    "customer_id": "count"
}).reset_index().rename(
    columns={"amount": "total_amount", "customer_id": "n_trans"})
ts_agg = ts_agg.sort_values("period")

fig, ax = plt.subplots(figsize=(11, 4))
ax.plot(ts_agg["period"], ts_agg["total_amount"], color="#2c7bb6", linewidth=1.5)
ax.fill_between(ts_agg["period"], ts_agg["total_amount"], alpha=0.2, color="#2c7bb6")
ax.yaxis.set_major_formatter(mticker.FuncFormatter(lambda x, _: f"${x/1e6:.1f}M"
                                                    if x >= 1e6 else f"${x/1e3:.0f}K"))
ax.set_title("Monthly Total Sales Revenue")
ax.set_xlabel("Month"); ax.set_ylabel("Revenue")
plt.tight_layout(); plt.show()
```

```{python ts-decomp-py}
ts_indexed = ts_agg.set_index("period")["total_amount"]

if len(ts_indexed) >= 2 * 12:
    decomp = seasonal_decompose(ts_indexed, model="additive", period=12)
    
    fig, axes = plt.subplots(4, 1, figsize=(11, 8))
    ts_indexed.plot(ax=axes[0], title="Original", color="#2c7bb6")
    axes[0].set_ylabel("Revenue")
    decomp.trend.plot(ax=axes[1], title="Trend", color="#1a9850")
    axes[1].set_ylabel("Trend")
    decomp.seasonal.plot(ax=axes[2], title="Seasonal", color="#fdae61")
    axes[2].set_ylabel("Seasonal")
    decomp.resid.plot(ax=axes[3], title="Residual", color="#8da0cb")
    axes[3].set_ylabel("Residual")
    plt.tight_layout(); plt.show()
```

:::

> **Interpretation:** Is there an upward or downward trend? What is the seasonal pattern? Example: "Trend is flat at ₦X per month; seasonality shows +12% lift in November–December (holiday season) and −8% dip in September (post-holiday discount period)."

---

## 10. Integrated Findings

> Write 200–250 words connecting the five analyses into one coherent narrative and a single prioritized recommendation.

---

## 11. Limitations & Further Work

- **Survey representativeness:** The 190 survey respondents may not represent the 8,017 sales transactions (survey response bias). Validate that survey respondents' purchasing patterns match the full customer distribution.
- **Customer matching:** Sales and survey data are linked by row number as a proxy; in production, a reliable customer ID or email match is essential to avoid spurious associations.
- **Temporal dynamics:** Clustering is static; in reality, customers move between segments over time (e.g., frequent buyers becoming dormant). A transition matrix or hidden Markov model would capture migration dynamics.
- **Revenue vs profit:** The analyses focus on revenue; profitability by customer segment may differ substantially due to cost-of-goods, acquisition cost, or logistics variation.
- **Future work:** Build a customer lifetime value (CLV) model integrating acquisition cost, retention probability, and repeat purchase rate to score customers on expected profitability over 12–36 months.

---

## References

> *(Complete in APA format)*

---

## Appendix: AI Usage Statement

> *(One paragraph describing AI assistance and independent judgement.)*
